---
title: "Supplementary Information: *Rtpca: an R package for differential thermal coaggregation analysis*"
author:
- name: Nils Kurzawa
  affiliation: 
  - European Molecular Biology Laboratory (EMBL), Genome Biology Unit
- name: André Mateus 
  affiliation: 
  - European Molecular Biology Laboratory (EMBL), Genome Biology Unit
- name: Mikhail M. Savitski
  affiliation: 
  - European Molecular Biology Laboratory (EMBL), Genome Biology Unit
date: "`r format(Sys.time(), '%d %B, %Y')`"
package: Rtpca
output:
  BiocStyle::pdf_document:
    keep_md: yes
vignette: >
    %\VignetteIndexEntry{Vignette Title}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8} 
bibliography: bibliography.bib
csl: cell.csl
header-includes: 
- \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Thermal proteome profiling (TPP) [@Savitski2014; @Mateus2020] is a mass spectrometry-based, proteome-wide implemention of the cellular thermal shift assay [@Molina2013]. It was originally developed to study drug-(off-)target engagement. However, it was realized that profiles of interacting protein pairs appeared more similar than by chance [@Tan2018, @Becher2018] which was coined as 'thermal proximity co-aggregation' (TPCA) [@Tan2018]. The R package `Rtpca` enables analysis of TPP datasets using the TPCA concept for studying protein-protein interactions and protein complexes and also allows to test for differential protein-protein interactions across different conditions.      
Here, we exemplify the analysis based on a dataset by @Becher2018 which provides temperature range TPP (TPP-TR) experiments for synchronized HeLa cells in G1/S cell cycle stage versus M phase.             
__Note:__ The paper by @Becher2018 also includes 2D-TPP [@Becher2016] data which is in general more sensitive to changes in protein abundance or stability. This data can also be informative on dynamics of protein-protein interactions based on correlations analysis of 2D-TPP profiles of annotated interactors. However, the advantage of TPP-TR data is that one can test for coaggregation which, if significant, is directly indicative of protein-protein interaction or complex assembly.

# Step-by-step walk through the data analysis

First, we need to load the required libraries (these need to be installed as specified in the comments):

```{r, message=FALSE, warning=FALSE}
library(dplyr) # install.packages("dplyr")
library(readxl) # install.packages("readxl")
library(Rtpca) # require(devtools); devtools::install_github("nkurzaw/Rtpca")
library(ggplot2) # install.packages("ggplot2")
```

Then, we download the supplementary data from Tan et al. which contains the TPP data which we'll be using:

```{r}
if(!file.exists("1-s2.0-S0092867418303854-mmc4.xlsx")){
    download.file(
        url = "https://ars.els-cdn.com/content/image/1-s2.0-S0092867418303854-mmc4.xlsx",
        destfile = "1-s2.0-S0092867418303854-mmc4.xlsx")
}

```
Next, we read in the annotation information of the supplementary table as a data frame

```{r}
supp_tab_becher_s4 <- read_xlsx("1-s2.0-S0092867418303854-mmc4.xlsx",
                                sheet = "TableS4_TPP-TR")

temperature_anno <- 
    as.numeric(
        gsub("T", "", gsub("_.+", "", colnames(
            supp_tab_becher_s4 %>% 
                dplyr::select(matches("mean\\.fc"))))))
```
We then extract the data for the 

```{r}
g1s_df <- supp_tab_becher_s4 %>% 
    filter(cell.cycle == "G1_S") %>% 
    dplyr::select(
        gene_name,
        replicates = found.in.reps,
        max_qupm = max.qupm,
        min_qupm = min.qupm,
        matches("mean\\.fc")) %>% 
    filter(min_qupm > 3,
           replicates == 3)
```

```{r}
g1s_mat <- as.matrix(
    g1s_df %>% dplyr::select(dplyr::matches("mean\\.fc"))
    )
rownames(g1s_mat) <- g1s_df$gene_name
attributes(g1s_mat)$temperature <- temperature_anno
```


```{r}
m_df <- supp_tab_becher_s4 %>% 
    filter(cell.cycle == "M") %>% 
    dplyr::select(
        gene_name,
        replicates = found.in.reps,
        max_qupm = max.qupm,
        min_qupm = min.qupm,
        matches("mean\\.fc")) %>% 
    filter(min_qupm > 3,
           replicates == 3)
```

```{r}
m_mat <- as.matrix(
    m_df %>% dplyr::select(dplyr::matches("mean\\.fc"))
    )
rownames(m_mat) <- m_df$gene_name
attributes(m_mat)$temperature <- temperature_anno
```


First, we load an annotation of mammalian complexes by @Ori2016, which comes with the `Rtpca` package:
```{r}
data("ori_et_al_complexes_df")
```


Then, we perform a TPCA anylsis only in the G1/S condition:

```{r}
G1S_TPCA <- runTPCA(
    objList = list(g1s_mat),
    complexAnno = ori_et_al_complexes_df)
```
We can plot the ROC curve for how predictive our data is on recovering protein complexes by evoking:

```{r}
plotComplexRoc(G1S_TPCA, computeAUC = TRUE)
```

And we can inspect significantly co-melting protein complexes, like this:
```{r}
G1S_TPCA@tpcaResultTable %>% filter(p_adj < 0.1)
```


```{r}
g1s_significant_complex_comelting <- 
    filter(G1S_TPCA@tpcaResultTable,p_adj < 0.1)$complex_name
```

Next, we perform the same analysis for only the M phase condition:

```{r}
M_TPCA <- runTPCA(
    objList = list(m_mat),
    complexAnno = ori_et_al_complexes_df)
```
We can see that the predictive performance of this dataset for protein complexes is not quite as good as for the G1/S one: 
```{r}
plotComplexRoc(M_TPCA, computeAUC = TRUE)
```

```{r}
M_TPCA@tpcaResultTable %>% filter(p_adj < 0.1)
```

Based on the protein complexes which we find significantly assembled in either condition, we will select the protein-protein interactions to test for in a differential TPCA:
```{r}
m_significant_complex_comelting <- 
    filter(M_TPCA@tpcaResultTable,p_adj < 0.1)$complex_name

all_significant_complex_comelting <- 
    unique(c(g1s_significant_complex_comelting, 
             m_significant_complex_comelting))
```

We load the annotation of protein-protein interactions within complexes that is composed of PPIs from StringDb [@Szklarczyk2019] and the complex annotation by @Ori2016 and filter it for protein complexes that we have seen to coaggregate in the analysis above.

```{r}
data("ori_et_al_complex_ppis")

filtered_complex_ppis <- ori_et_al_complex_ppis %>% 
    filter(complex_name %in% all_significant_complex_comelting) 
```

We now run the differential TPCA by evoking""
```{r}
set.seed(123)
M_vs_G1S_diff_TPCA <- runDiffTPCA(
    objList = list(g1s_mat),
    contrastList = list(m_mat),
    ctrlCondName = "G1/S",
    contrastCondName = "M",
    ppiAnno = filtered_complex_ppis, 
    n = 10^6
)
```

We can now plot the result in form of a volcano plot:

```{r}
plotDiffTpcaVolcano(M_vs_G1S_diff_TPCA, 
                    setXLim = TRUE,
                    xlimit = c(-0.75, 0.75))
```

We can now inspect the significant results:
```{r}
M_vs_G1S_diff_TPCA@diffTpcaResultTable %>% 
    dplyr::select(pair, rssC1_rssC2, p_value, p_adj) %>% 
    arrange(p_value)
```

To validate significant PPIs we can inspect their melting curves:
```{r}
plotPPiProfiles(M_vs_G1S_diff_TPCA, pair = c("NUP205", "RANBP2"))
```

```{r}
plotPPiProfiles(M_vs_G1S_diff_TPCA, pair = c("RPS6", "RPSA"))
```


```{r}
sessionInfo()
```

# References

