---
title: "Supplementary Information: *Rtpca: an R package for differential thermal coaggregation analysis*"
author:
- name: Nils Kurzawa
  affiliation: 
  - European Molecular Biology Laboratory (EMBL), Genome Biology Unit
- name: André Mateus 
  affiliation: 
  - European Molecular Biology Laboratory (EMBL), Genome Biology Unit
- name: Mikhail M. Savitski
  affiliation: 
  - European Molecular Biology Laboratory (EMBL), Genome Biology Unit
date: "`r format(Sys.time(), '%d %B, %Y')`"
package: Rtpca
output:
  BiocStyle::pdf_document:
    keep_md: yes
vignette: >
    %\VignetteIndexEntry{Vignette Title}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8} 
bibliography: bibliography.bib
csl: cell.csl
header-includes: 
- \usepackage{placeins}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Thermal proteome profiling (TPP) [@Savitski2014; @Mateus2020] is a mass spectrometry-based, proteom-wide implemention of the cellular thermal shift assay [@Molina2013]. It was originally developed to study drug-(off-)target engagement. However, it was realized that profiles of interacting protein pairs appeared more similar than by change [@Tan2018, @Becher2018] which was coined as 'thermal proximity co-aggregation' (TPCA) [@Tan2018]. The R package `Rtpca` enables analysis of TPP datasets in the light of protein-protein interactions or protein complex associations. Here, we exemplify the analysis based on a dataset by Hashimoto et al. [@Hashimoto2020] studying the dynamics of protein-protein interactions in human cells during Cytomegalovirus infection.

# Step-by-step walk through the data analysis

First, we need to load the required libraries (these need to be installed as specified in the comments):

```{r, message=FALSE, warning=FALSE}
library(dplyr) # install.packages("dplyr")
library(readxl) # install.packages("readxl")
library(Rtpca) # require(devtools); devtools::install_github("nkurzaw/Rtpca")
library(ggplot2) # install.packages("ggplot2")
```


Then, we download the supplementary data from Hashimoto et al. which contains the TPP data:

```{r}
if(!file.exists("41467_2020_14586_MOESM4_ESM.xlsx")){
    download.file(
    url = "https://static-content.springer.com/esm/art%3A10.1038%2Fs41467-020-14586-5/MediaObjects/41467_2020_14586_MOESM4_ESM.xlsx", 
    destfile = "41467_2020_14586_MOESM4_ESM.xlsx")
}

```
We then read in the excel table:
```{r}
suppl1_df <- read_xlsx("41467_2020_14586_MOESM4_ESM.xlsx", sheet = "TableS1B") 
```

From the table we create a matrix of the median relative fold changes for the Mock dataset:
```{r}
mock_df <- suppl1_df %>% 
    dplyr::select(Uniprot_ID, Gene_symbol, dplyr::matches("Mock")) %>% 
    na.omit()
mock_mat <- as.matrix(mock_df[,-c(1,2)])
rownames(mock_mat) <- mock_df$Gene_symbol
id_dup <- which(duplicated(rownames(mock_mat)))
mock_mat <- mock_mat[-id_dup,]
```

The same is done for the 24h post infection data:
```{r}
inf24h_df <- suppl1_df %>% 
    dplyr::select(Uniprot_ID, Gene_symbol, dplyr::matches("24")) %>% 
    na.omit()
inf24h_mat <- as.matrix(inf24h_df[,-c(1,2)])
rownames(inf24h_mat) <- inf24h_df$Gene_symbol
id_dup_24h <- which(duplicated(rownames(inf24h_mat)))
inf24h_mat <- inf24h_mat[-id_dup_24h,]
```


We now load our annotation of protein pairs within the same protein complexes (this dataset comes with the `Rtpca` package, please make sure to cite it whenever you use it @Ori2016):
```{r}
data("ori_et_al_complex_ppis")
```


We extend this annotation by one of the significantly changing protein-protein interactions that Hashimoto et al. reported that is not present in our annotation:
```{r}
ori_et_al_complex_ppis_extended <- 
    bind_rows(ori_et_al_complex_ppis,
           tibble(
               complex_name = "PI3K Pathway",
               x = "CRK",
               y = "PIK3R1",
               pair = "CRK:PIK3R1"))
```

## Running Rtpca for a single condition

To run Rtpca for the mock dataset:
```{r cache=TRUE}
mockTPCA <- runTPCA(
    objList = list(mock_mat),
    ppiAnno = ori_et_al_complex_ppis
)
```
We can then look at the ROC curve with:
```{r cache=TRUE, fig.cap="ROC curve for identification of protein pairs part of the same complex in the Mock dataset."}
plotPPiRoc(mockTPCA, computeAUC = TRUE)
```
\FloatBarrier
and the result table of significantly co-melting protein pairs:
```{r cache=TRUE}
head(mockTPCA@tpcaResultTable %>% filter(p_adj < 0.01) %>% arrange(p_adj))
```
The same can be done for the 24h post infection dataset:
```{r cache=TRUE}
inf24hTPCA <- runTPCA(
    objList = list(inf24h_mat),
    ppiAnno = ori_et_al_complex_ppis
)
```

```{r cache=TRUE, fig.cap="ROC curve for identification of protein pairs part of the same complex in the 24h post infection dataset."}
plotPPiRoc(inf24hTPCA, computeAUC = TRUE)
```
\FloatBarrier
```{r}
head(inf24hTPCA@tpcaResultTable %>% filter(p_adj < 0.01) %>% arrange(p_adj))
```

## Running a differential Rtpca analysis

To run the differential TPCA test between the mock and the 24h post infection datasets we run:
```{r cache=TRUE}
mockInf24hDiffTPCA <- runDiffTPCA(
    objList = list(mock_mat),
    contrastList = list(inf24h_mat),
    ppiAnno = ori_et_al_complex_ppis_extended
)
```

And we can inspect the result with a volcano plot:
```{r fig.cap="Volcanoplot of differential TPCA results. Colored in black are the significantly differentially coaggregating protein pairs betwee both datasets."}
plotDiffTpcaVolcano(mockInf24hDiffTPCA)
```
\FloatBarrier
The interaction between CRK and PIK3R1 is one of the significantly changing interations, which we can see by labeling it in the volcano plot:
```{r fig.cap="Volcanoplot of differential TPCA results. Colored in red is the protein pair CRK-PIK3R1 which is significantly coaggregating more strongly in the 24h post infection data."}
plotDiffTpcaVolcano(mockInf24hDiffTPCA) +
    geom_point(color = "red",
               data = filter(mockInf24hDiffTPCA@diffTpcaResultTable, 
                             pair == "CRK:PIK3R1"))
```
\FloatBarrier
As we can see from figure \@ref(fig:riboVolcanoPlot) many of the protein-protein interactions that appear to losen up in the post infection dataset compared to Mock are between members of ribosome. This could reflect translational stalling in response to viral infection.

```{r riboVolcanoPlot, cache=TRUE, fig.cap="Volcanoplot of differential TPCA results. Colored in blue are significantly changing protein pairs including at least one ribosomal protein."}
plotDiffTpcaVolcano(mockInf24hDiffTPCA) + 
    geom_point(color = "steelblue",
               data = filter(mockInf24hDiffTPCA@diffTpcaResultTable, 
                             p_adj < 0.1,
                             grepl("^RP[L,S]", pair)))
```


```{r}
sessionInfo()
```

# References